
\chapter{Case studies}
\label{casestudies}

\section{Case Study 1: Michael Giacchino}

Michael Giacchino (1967). website:\url{http://www.michaelgiacchinomusic.com/}. Note well his dedication to his education. He received a very thorough grounding in film, art and music. His major career boost came when he began writing for games including the well known \textit{Medal of Honor} series. He then teamed up with J.J.Abrams and wrote for a number of TV series. And through this collaboration he wrote for \textit{Star Trek} and \textit{Star Trek Into Darkness}.

His collaboration with Brad Bird have included \textit{The Incredibles}, \textit{Ratatouille}. His collaboration with Matt Reeves resulted in a score for \textit{Dawn of the Planet of the Apes}. His collaboration with Pete Doctor resulted in a score for \textit{UP}. His collaboration with Andrew Stanton resulted in a score for \textit{John Carter}.

\subsection{John Carter}
\begin{itemize}
\item Science fiction but with a Wild West slant.   
\item Horn theme. A lot of music backing the picture from simple drones to pulsing percussion.  
\item A three note theme. Open fifth (chord I to IVmin)
\item Use of Ostinati and heavy reliance upon strings.
\item Similarities with Star Trek especially in the love theme (around 1:15:00 and 1:17:30)
\end{itemize}

\section{Case Study 2: Bernard Hermann}
\begin{itemize}
\item \href{http://www.bernardherrmann.org/}
{The Society for the Appreciation of the Music of Bernard Hermann}
\item
\item
\end{itemize}

\section{Case Study 3: Stanley Kubrick}
Ligeti's music is closely linked to the films of Stanley Kubrick. 
\begin{itemize}
\item 2001: A Space Odyssey
\begin{itemize}
\item Atmosphères (StarGate sequence and other places) 
\item Lux Aeterna (for the moon-bus scene en route to the TMA-1 monolith)
\item Requiem (Kyrie linked to the black monolith)
\item Aventures (closing scenes)
\item There are small recurrences in the film 2010.
\end{itemize}
\item The Shining uses small portions of Lontano for orchestra alongside music in a similar vein by Penderecki (Utrenja, De Natura Sonoris No.1 and No.2, and Polymorphia)
\end{itemize}


